\mutagen: Working title

### 2021-03-29 papers/oopsla-21/main.tex
WIP

The next section introduces three heurisitics we added to the base testing loop
of \mutagen based on the limitations we found in \fuzzchick.
%
These heuristics can that offer a substantial improvement in practice.

@@

Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies. Integer placerat tristique nisl. Praesent augue. Fusce
commodo.

@@

Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
vitae placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis,
varius et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

@@

\begin{algorithm}[H]
  \SetArgSty{textnormal}
  \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    $\cdots$\;
    (result, trace) $\gets$ WithTrace(P(x))\;
    $\cdots$\;
    \If{Interesting(TLog, trace)}{
      \If{\KwNot Discarded(result)}{
        muts $\gets$ Mutants(x, mut, R)\;
        {\color{red} prio $\gets$ BranchDepth(TLog, trace)}\; %% NEW
        {\color{red} PushFront(QSucc, prio, muts)}\;          %% NEW
      }
      $\cdots$\;
    }
  }
  \vspace{10pt}
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QSucc, QDisc, gen}}{
    \If{\KwNot Empty(QSucc)}{
      {\color{red} (muts, prio) $\gets$ DequeMin(QSucc)}\; %%NEW
      \If{Empty(muts)}{
        Pick(QSucc, QDisc, gen)\;
      } \uElse{
        {\color{red} PushFront(QSucc, prio, Rest(muts))}\; %% NEW
        \KwRet First(muts)\;
      }
    }
    $\cdots$\;
  }

\caption{Priority FIFO Heuristic}
\end{algorithm}

@@

\begin{algorithm}[H]
  \SetArgSty{textnormal}
  \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, gen, mut}}{
    {\color{red}
    boring  $\gets$ 0\;
    reset   $\gets$ 1000\;
    R       $\gets$ 1\;
    }
    $\cdots$\;
    \While{i $<$ N}{
      {\color{red}
      \If{boring $>$ reset}{
        TLog   $\gets$ $\varnothing$\;
        reset $\gets$ reset * 2\;
        R     $\gets$ R * 2\;
      }
      }
      $\cdots$\;
      \lIf{\KwNot result}{\KwRet Bug(x)}
      \If{Interesting(TLog, trace)}{
        $\cdots$\;
      }
      \lElse {
        {\color{red} boring $\gets$ boring + 1}
      }
      $\cdots$\;
    }
  }

\caption{Trace Saturation Heuristic}
\end{algorithm}

\end{multicols}
\caption{\label{fig:heuristics} The two main heuristics implemented in \mutagen.
  Statements in {\color{red} red} indicate important changes to the base
  algorithm. Ellipses denote parts of the code that are not relevant for the
  point being made. }
\end{figure}

@@

Vestibulum convallis, lorem a tempus semper, dui dui euismod elit, vitae
placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis, varius
et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat tellus ut
neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus lacinia purus,
et dictum nunc justo sit amet elit.

\subsection{Mutation Inheritance}

Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
vitae placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis,
varius et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

@@

\section{Implementation}
\label{sec:implementation}

\subsection{Deriving Mutation Machinery}

Integer placerat tristique nisl. Praesent augue. Fusce commodo. Vestibulum
convallis, lorem a tempus semper, dui dui euismod elit, vitae placerat urna
tortor vitae lacus. Nullam libero mauris, consequat quis, varius et, dictum id,
arcu. Mauris mollis tincidunt felis. Aliquam feugiat tellus ut neque. Nulla
facilisis, risus a rhoncus fermentum, tellus tellus lacinia purus, et dictum
nunc justo sit amet elit.

\subsection{Tracing Haskell Programs}

\begin{verbatim}
  sorted []       = True
  sorted [x]      = True
  sorted (x:y:xs) = if x <= y then sorted (y:xs) else False
\end{verbatim}

\begin{verbatim}
  sorted []       = _trace_ 1 (True)
  sorted [x]      = _trace_ 2 (True)
  sorted (x:y:xs) = _trace_ 3 (if x <= y then _trace_ 4 (sorted (y:xs)) else _trace_ 5 (False))
\end{verbatim}

Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies.

>>>>

The next section introduces two heurisitics we added to the base testing loop of
\mutagen based on the limitations we found in \fuzzchick.
%
%% These heuristics can that offer a substantial improvement in practice.

@@

In this section we introduce two heuristics implemented on top of the base
testing loop of our tool.
%
\mutagen enables them all by default, although they can be individually disabled
by the user if deemed appropriate.

@@

This heuristic tackles the issue of enqueuing novel mutation candidates at the
end of (possibly) long queues of not-so-interesting previously executed ones.

\fuzzchick uses AFL instrumentation under the hood, which in turn uses an
\emph{edge coverage} criteria to distinguish novel executions and to assign each
mutation candidate a given energy.
%
In contrast, execution traces \mutagen represent the \emph{path} in the code
taken by the program, as opposed to just the set of edges traversed in the
control-flow graph (CFG).
%
Using this criteria lets us gather precise information from the each new
execution.
%
%% To keep track of the execution path of each test, our tool keeps an internal
%% prefix tree that records every new entry by traversing this data structure and
%% inserting new nodes after a new path prefix gets discovered.
%% %
In particular, we are interested in the \emph{depth} where each new execution
branches from already seen ones.
%
Our assumption here is that test cases that differ (branch) at shallower depths
from the ones already executed are more likely to discover completely new
portions of the code under test, and hence we want to assign them a higher
priority.

In this light, every time we insert a new execution path into the internal trace
log, we calculate the number of new nodes that were executed, as well as the
\emph{branching depth} where they got inserted.
%
The former is used to distinguish interesting test cases (whether or not new
nodes were inserted), whereas the latter is used by this heuristic to schedule
mutation candidates.
%
Fig. \ref{fig:mutagen:tracelog} illustrates this idea, inserting two execution
traces (one after another) into a trace log that initially contains a single
execution path.
%
The second insertion (with trace $1 \rightarrow 2 \rightarrow 6 \rightarrow 7$)
branches at a shallower depth than the first one (2 vs. 3), hence its
corresponding test case should be given a higher priority.

\begin{figure}[b]
\newcommand\stacked[2]{\underset{\textstyle #2\mathstrut}{#1}}
\begin{equation}
  \nonumber
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\color{red} \small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\color{blue} \small \texttt{5}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=1\phantom{x}}}{{\color{red} depth=3}}
  \qquad
  \qquad
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
      }
  }
  \right]
  \;
  +
  \;
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=0.75cm}
  ]{
    \node {\small \texttt{1}}
      child {node {\small \texttt{2}}
        child {node {\small \texttt{6}}
          child {node {\small \texttt{7}}
          }
        }
      }
  }
  \;
  =
  \;
  \left[
  \tikz[
    baseline=-10mm,
    level distance=0.6cm,
    level 1/.style={sibling distance=0.75cm},
    level 2/.style={sibling distance=1cm}
  ]{
    \node {\color{red} \small \texttt{1}}
      child {node {\color{red} \small \texttt{2}}
        child {node {\small \texttt{3}}
          child {node {\small \texttt{4}}}
          child {node {\small \texttt{5}}}
        }
        child {node {\color{blue}\small \texttt{6}}
          child {node {\color{blue}\small \texttt{7}}}
        }
      }
  }
  \right]
  ,
  \stacked{{\color{blue} new=2\phantom{x}}}{{\color{red} depth=2}}
\end{equation}
\caption{\label{fig:mutagen:tracelog}Inserting two new execution traces into the
  internal trace log (represented using brackets).}
\end{figure}

With this mechanism in place, we can modify \mutagen's base testing loop by
replacing each mutation queue with a priority queue indexed by the branching
depth of each new execution.
%
These changes are illustrated in Algorithm \ref{algo:mutagen:fifo}.
%
Statements in {\color{red} red} indicate important changes to the base
algorithm, whereas ellipses denote parts of the code that are not relevant for
the point being made.

To pick the next test case, we simply retrieve the one with highest priority
(the smallest branching depth).
%
Then, whenever we find a new interesting test case, we enqueue it at the
beginning of the queue of its corresponding priority.
%
This allows the testing loop to jump immediately onto processing new interesting
candidates as soon as they are found (even at the same priority), and to jump
back to previous candidates as soon as their mutants become progressively less
novel.

@@

\begin{algorithm}[H]
  \SetArgSty{textnormal}
  \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    $\cdots$\;
    x $\gets$ Pick(QSucc, QDisc, gen, mut)\;
    (result, trace) $\gets$ WithTrace(P(x))\;
    $\cdots$\;
    \If{Interesting(TLog, trace)}{
      \If{\KwNot Discarded(result)}{
        muts $\gets$ Mutants(x, mut, R)\;
        {\color{red} prio $\gets$ BranchDepth(TLog, trace)}\; %% NEW
        {\color{red} PushFront(QSucc, prio, muts)}\;          %% NEW
      }
      $\cdots$\;
    }
  }
  \vspace{5pt}
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QSucc, QDisc, gen}}{
    \If{\KwNot Empty(QSucc)}{
      {\color{red} (muts, prio) $\gets$ DequeMin(QSucc)}\; %%NEW
      \If{Empty(muts)}{
        Pick(QSucc, QDisc, gen)\;
      } \uElse{
        {\color{red} PushFront(QSucc, prio, Rest(muts))}\; %% NEW
        \KwRet First(muts)\;
      }
    }
    $\cdots$\;
  }

\caption{\label{algo:mutagen:fifo}Priority FIFO Heuristic}
\end{algorithm}

@@

\begin{algorithm}[H]
  \SetArgSty{textnormal}
  \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, gen, mut}}{
    {\color{red}
    boring  $\gets$ 0\;
    reset   $\gets$ 1000\;
    R       $\gets$ 1\;
    }
    $\cdots$\;
    \While{i $<$ N}{
      {\color{red}
      \If{boring $>$ reset}{
        TLog   $\gets$ $\varnothing$\;
        reset $\gets$ reset * 2\;
        R     $\gets$ R * 2\;
      }
      }
      $\cdots$\;
      \lIf{\KwNot result}{\KwRet Bug(x)}
      \If{Interesting(TLog, trace)}{
        {\color{red} boring  $\gets$ 0}\;
        $\cdots$\;
      }
      \lElse {
        {\color{red} boring $\gets$ boring + 1}
      }
      $\cdots$\;
    }
  }

\caption{\label{algo:mutagen:reset}Trace Saturation Heuristic}
\end{algorithm}

\end{multicols}
%% \caption{\label{fig:heuristics} The two main heuristics implemented in \mutagen.
%%   Statements in {\color{red} red} indicate important changes to the base
%%   algorithm. Ellipses denote parts of the code that are not relevant for the
%%   point being made. }
\end{figure}

@@

As introduced in Section \ref{sec:mutagen}, our tool is parameterized by the
amount of random mutations to be generated over each mutable subexpression
defined using a random mutant, e.g., numeric values, characters, etc.
%
But, how many random mutations should we use? A single one? A few tenths? A few
hundreds?
%
Clearly, using too little, can put finding bugs at risk.
%
For instance, when the system under test branches on numeric values, we should
make sure that we set enough random mutations to test each case in a reasonable
basis.
%
Using too many, the other hand, can degrade the performance of the testing
loop, as it will spend too much time producing uninteresting mutations.
%
This can happen for instance if the subexpressions defined using random mutants
are only used as payloads, and their value does not affect the execution in any
way.

Answering this question precisely is not an easy task, and the second heuristic
we introduce in this work aims to tackle this issue.
%
We found that, the smaller the number of random mutations we set, the easier it
is for the trace log that records executions to start getting saturated, i.e.,
when interesting test cases stop getting discovered or are discovered very
sporadically.
%
Our realization is that we can use this information to automatically optimize
the number of random mutations used by our tool.
%
This idea is described in Algorithm \ref{algo:mutagen:reset}.
%
The process is simple:
%
\begin{inparaenum}
  \item we start the testing loop with the amount of random mutations set to one,
  %
  \item each time we find that a test is not interesting (i.e. boring), we
    increment a counter,
  \item if we have not produced any interesting test case after a certain
    threshold (1000 tests seems to be a reasonable value in practice), we
    increment the amount of random mutations and the threshold by twice the
    current amount.
    %
    Additionally, each time this happens we also reset the trace log, so
    interesting test cases found on a previous iteration can be found and
    enqueued for mutation again --- this time with a higher effort dedicated to
    producing random mutations.
\end{inparaenum}

Then, if the execution of the system under test depends heavily on the values
stored at randomly mutable subexpressions, starting with a single random mutation
will quickly saturate the trace space, and this heuristic will continuosly
increase the random mutations parameter until that stops happening.

%% \subsection{Mutation Inheritance}

%% The last heuristic we discuss in this work attempts to reduce the amount of
%% mutations required when testing properties that lazily consume foldable inputs
%% like lists or trees.
%% %
%% To illustrate this, suppose we define a property that aims to test a list
%% sorting algorithm:

@@

%% \section{Implementation}
%% \label{sec:implementation}

%% \subsection{Deriving Mutation Machinery}

%% Integer placerat tristique nisl. Praesent augue. Fusce commodo. Vestibulum
%% convallis, lorem a tempus semper, dui dui euismod elit, vitae placerat urna
%% tortor vitae lacus. Nullam libero mauris, consequat quis, varius et, dictum id,
%% arcu. Mauris mollis tincidunt felis. Aliquam feugiat tellus ut neque. Nulla
%% facilisis, risus a rhoncus fermentum, tellus tellus lacinia purus, et dictum
%% nunc justo sit amet elit.

%% \subsection{Tracing Haskell Programs}

%% \begin{verbatim}
%%   sorted []       = True
%%   sorted [x]      = True
%%   sorted (x:y:xs) = if x <= y then sorted (y:xs) else False
%% \end{verbatim}

%% \begin{verbatim}
%%   sorted []       = _trace_ 1 (True)
%%   sorted [x]      = _trace_ 2 (True)
%%   sorted (x:y:xs) = _trace_ 3 (if x <= y then _trace_ 4 (sorted (y:xs)) else _trace_ 5 (False))
%% \end{verbatim}

%% Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
%% tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
%% leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
%% wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
%% mollis ultricies.
